---
title: "Translating Stata to R"
output:
  bookdown::html_document2: 
    number_sections: false 
    toc: true
    toc_float: true
    smart: false
---


```{r, include=FALSE}
tutorial::go_interactive(greedy=FALSE,  height = 290)
```

# Index {#index}

This manual explains the R syntax for people familiar with Stata.<br>
NOTE: the examples only illustrate R syntax without considering specifically how the data should be analysed to take into account its particular properties.<br><br>

[about](#about) \ \  [ameans](#ameans) \ \  [anova](#anova) \ \  [append](#append) <br>
[bootstrap](#bootstrap) \ \  [browse](#browse) \ \  [bsample](#bsample) \ \  [by](#by) \ \  [bysort](#by) <br>
[cd](#cd) \ \  [centile](#centile) \ \  [clear](#clear) \ \  [clogit](#clogit) \ \  [cls](#clear) \ \  [codebook](#summarize) \ \  [collapse](#collapse) \ \  [correlate](#correlate) \ \  [count](#count) <br>
[describe](#describe) \ \ [dotplot](#dotplot) \ \  [drop](#drop) <br>
[egen](#egen) \ \  [estimates_store](#lrtest) <br>
[findit](#help) <br>
[generate](#generate) \ \  [graph_bar](#bar) \ \  [graph_box](#box) \ \  [graph_matrix](#matrix) <br>
[help](#help) \ \  [histogram](#hist)<br>
[if](#if) \ \  [in](#in) <br>
[kappa](#kappa) \ \  [keep](#drop) \ \  [ktau](#correlate)<br> 
[levelsof](#levelsof) \ \  [list](#list) \ \  [logit](#logit) \ \  [logistic](#logit) \ \  [lowess](#lowess) \ \  [lrtest](#lrtest) <br>
[mean](#mean) \ \  [merge](#merge) \ \  [mhodds](#mhodds) \ \  [mvdecode](#mvdecode)\ \  [mvenecode](#mvdecode) <br>
[predict](#predict) <br>
[qnorm](#qnorm) <br>
[ranksum](#ttest) \ \  [recode](#recode) \ \  [regress](#regress) \ \  [rename](#rename) \ \  [rvfplot](#regress) \ \  [replace](#replace) \ \  [reshape](#reshape) <br>
[sample](#sample) \ \  [save](#save) \ \  [scatter](#scatter) \ \  [signrank](#ttest) \ \  [sort](#sort) \ \  [spearman](#correlate) \ \  [summarize](#summarize) <br> 
[tab1](#tabulate) \ \  [table](#by) \ \  [tabstat](#collapse) \ \  [tabulate](#tabulate) \ \ [ttest](#ttest) \ \  [twoway_lfit](#lowess)<br> 
 [xi](#xi) \ \ \  [xtile](#xtile)<br><br> 



# Example dataset <tt>airquality</tt>

Data represent daily air quality measurements in New York, May to September 1973.
The variables <tt>Ozone</tt> and <tt>Solar.R</tt> have several missing values.<br>  
NOTE: the examples only illustrate R syntax without considering specifically how the data should be analysed to take into account its particular properties.<br><br>  

```{r tut=F, echo=F}
knitr::kable(data.frame(No = 1:6, 
        Var =  colnames(airquality), 
        Type=rep("number",6),
        Description =c(
          "Ozone (ppb)",
          "Solar radiation (lang)",
          "Wind speed (mph)",
          "Temperature (°F)",
          "Month (1--12)",
          "Day of month (1--31)"
    )), booktabs = TRUE,
  caption = 'Example dataset airquality'
)

```

<br><br>Return to [index](#index)<br>


# Example dataset <tt>infert</tt>
Matched case-control study (ratio 1:2) on infertility after spontaneous and induced abortion.
One case with two prior spontaneous abortions and two prior induced abortions is omitted.
The first variable <tt>education</tt> is a categorical variable - called a factor in R.<br>  
NOTE: the examples only illustrate R syntax without considering specifically how the data should be analysed to take into account its particular properties.<br>   
```{r tut=F, echo=F}
knitr::kable(data.frame(No = 1:8, Var = 
        colnames(infert), 
        Type=c("category",rep("number", 3),"binary",rep("number", 3)),
        Description =c(
          "Years of eduction 0:<6; 1:6-11; 2:12+",
          "Age (years)",
          "Parity",
          "Numebr of prior induced abortions",
          "Case (1) or control (0)",
          "Numebr of prior spontaneous abortions",
          "Matching ID",
          "Pooled matching ID"
          )), booktabs = TRUE,
  caption = 'Example dataset infert'
)
```

<br>Return to [index](#index)<br>


# about {#about}

The equivalent function in R is <tt>sessionInfo()</tt><br>   
<br>Return to [index](#index)<br>

# ameans {#ameans}

The basic function for the arithmetic mean is <tt>mean</tt>. 
Please note that this function will return NA if any value is missing.
To ignore missing values add the argument <tt>na.rm = TRUE</tt>
R has no inbuild function for calculation the geometric mean.
However, it is easy to define a function.
Keep in mind: The geometric mean is 0 if any of the values is 0.
In case of 0's a constant is often added to each value (usually 1).<br>
The mean for all variables is easily obtained with <tt>[summary](#summarize)</tt>.<br><br>

```{r tut=TRUE}
# STATA ameans Solar.R
# arithmetic mean
mean(airquality$Solar.R)
# NOTE: mean will return NA if any value is missing
#       Argument 'na.rm=T' ignores missing values 
mean(airquality$Solar.R, na.rm=T)
# geometric mean (no missing)
geomean <- function(x) exp(sum(log(x))/length(x))
geomean(airquality$Temp)
# STATA ameans Temp, add(1)
geoadd1 <- function(x) exp(sum(log(x+1))/length(x))
geoadd1(airquality$Temp)

```
<br>Return to [index](#index)<br>

# anova {#anova}

R interprets ANOVA as a special case of linear regression.
The function which shows the Analysis of Variance table is <tt>anova</tt> 
which is called on a regression object.<br><br> 

```{r tut=TRUE}
# Binary predictor
# STATA anova age case  
anova(lm(age ~ case, data=infert))
# Categorical predictor - categorical (factor) variable in R
# STATA xi: anova age i.education 
anova(lm(age ~ education, data=infert))
# Categorical predictor - numerical variable in R
# STATA xi: anova age i.induced 
anova(lm(age ~ factor(induced), data=infert))

```
<br>Return to [index](#index)<br>

# append {#append}

To append two datasets, R has the function <tt>rbind</tt>  (for 'row bind').
Note: Both datasets must have the same number of variables with identical variable names.
However, the order of variables may be different.<br><br> 

```{r tut=TRUE}
# Make a copy of dataset infert
inf.copy <- infert
# Append both datasets 
# STATA append using ...  
new.data <- rbind(infert, inf.copy)
nrow(infert)
nrow(new.data)

```
<br>Return to [index](#index)<br>

# bootstrap {#bootstrap}

A bootstrap algorithm is implemented in the package <tt>boot</tt>.
See [centile](#centile) for an example how to estimate confidence intervals for a quantile.<br><br>

```{r tut=TRUE}
# 95% CI of the mean via bootstrap
# STATA bootstrap r(mean), reps(1000): summarize Temp
library(boot)
mean.boot <- boot(data = airquality$Temp, statistic = function(x, i) mean(x[i]), R = 1000)
mean.boot 
boot.ci(mean.boot)

```
<br>Return to [index](#index)<br>

# browse {#browse}

In R: <tt>View</tt>.

<br>
<tt>
View(infert)
</tt>
<br>
<br>Return to [index](#index)<br>

# bsample {#bsample}

For Stata's (bootstrap) sample with replacement use  <tt>sample</tt> with argument <tt>replace = TRUE</tt>. <br><br> 

```{r tut=TRUE}
# Sampling the numbers from 1 to 20 with replacement
sample(1:20, replace = T)
# ... sorted:
sort( sample(1:20, replace = T))
# STATA bsample Temp
sample(airquality$Temp, replace = T)

```
<br>Return to [index](#index)<br>

# by bysort table {#by}

R does not require the dataset to be sorted. Therefore, bysort is not needed in R.
The syntax of R's equivalent function is <tt>by(values, bygroups, function)</tt>.
R's <tt>by</tt> is also a useful alternative to Stata's <tt>table</tt>.<br><br> 

```{r tut=TRUE}
# STATA by Month: mean Solar.R
by(airquality$Solar.R, airquality$Month, mean)

# NOTE: mean will return NA if any value is missing
#       Argument 'na.rm=T' ignores missing values 
by(airquality$Solar.R, airquality$Month, mean, na.rm=T)

# STATA table Month,  c(min Solar.R mean mpg median Solar.R max Solar.R): mean Solar.R
by(airquality$Solar.R, airquality$Month, summary)

```
<br>Return to [index](#index)<br>

# cd {#cd}

R's function to change the directory is <tt>setwd</tt> (for 'set working directory').
Note: R understands only forward slash! If the path contains back slashes (typically for Windows) you have to convert them to forward slashes. <br>
Examples:<br><br>
<tt>
setwd("C:/Users/KM/Documents/R/")
</tt>

<br><br>Return to [index](#index)<br>

# centile {#centile}

Percentiles of a single variable are provided by <tt>quantile</tt>.
Please note that this function will return NA if any value is missing.
To ignore missing values add the argument <tt>na.rm = TRUE</tt>.
The median, 1st quantile and 3rd quantile  for all variables is easily obtained with <tt>[summary](#summarize)</tt>.
The confidence interval of the pseudo median can be obtained with <tt>wilcox.test</tt>.
For confidence interval of other percentiles [bootstrap](#bootstrap) resampling is recommended.<br><br>

```{r tut=TRUE}
# STATA centile
summary(airquality)
# obtain 95%CI of the median
# STATA centile Temp
wilcox.test(airquality$Temp, conf.int = T)
# STATA centile Ozone, centile(10, 50, 90) 
quantile(airquality$Ozone, probs=c(0.1, 0.5, 0.9), na.rm=T)
# 95% CI of the 1st quantile via bootstrap
library(boot)
q1.boot <- boot(data = airquality$Temp, statistic = function(x, i) quantile(x[i], probs=.25), R = 1000)
q1.boot 
boot.ci(q1.boot)

```
<br>Return to [index](#index)<br>



# clear cls {#clear}

To remove a certain dataset  (or any R object) from memory use <tt>rm(mydata)</tt>.
To remove all objects use <tt>rm(list=ls())</tt>.
To clear the text in the console press [CTRL] + [L].

<br><br>Return to [index](#index)<br>


# clogit {#clogit}

Package <tt>survival</tt> provides an equivalent function for conditional logistic regression.
The model formula is of the form <tt>case.status ~ exposure + strata(matched.set)</tt>. 
<br><br>

```{r tut=TRUE}
# STATA clogit case spontaneous induced, group(stratum)
library(survival)
clog <- clogit(case ~ spontaneous + induced + strata(stratum), data=infert)
summary(clog)

```
<br>Return to [index](#index)<br>


# codebook summarize {#summarize}

In R <tt>summary</tt> provides many information on all or a few variables.
For percentiles see [centile](#centile). <br><br>

```{r tut=TRUE}
# STATA summarize
summary(airquality)

# Number of unique values of Day
# STATA codebook Day
sort(unique(airquality$Day))
length(unique(airquality$Day))

```
<br>Return to [index](#index)<br>


# collapse tabstat {#collapse}

Package <tt>dplyr</tt> provides several functions to generate summary statistics for numeric variables.<br><br> 

```{r tut=TRUE}
# STATA tabstat Temp Ozone, by(Month) stat(mean median n)
library(dplyr)
airquality %>%
  group_by(Month) %>%
  summarize(mean.Te = mean(Temp, na.rm = TRUE),
            median.Te = median(Temp, na.rm = TRUE),
            mean.Oz = mean(Ozone, na.rm = TRUE),
            median.Oz = mean(Ozone, na.rm = TRUE),
            n =n())
```
<br>Return to [index](#index)<br>


# correlate ktau spearman {#correlate}

Use <tt>cor</tt> in R.<br><br>

```{r tut=TRUE}
# STATA correlate _all
cor(airquality,  use = "pairwise.complete.obs")

# STATA spearman _all
cor(airquality,  use = "pairwise.complete.obs", method = "spearman")

# STATA ktau _all
cor(airquality,  use = "pairwise.complete.obs", method = "kendall")

```
<br>Return to [index](#index)<br>


# count {#count}

The number of records or the number fullfilling a certain conditions can be 
determined with <tt>nrow</tt>.<br><br> 

```{r tut=TRUE}
# STATA count
nrow(airquality)
# STATA count if month < 9
nrow(airquality[airquality$Month < 8, ])
#with subset
nrow(subset(airquality, Month < 8))
#with package dplyr
library(dplyr)
filter(airquality, Month < 8) %>% nrow()

```
<br>Return to [index](#index)<br>


# describe {#describe}

<tt>str</tt> (for 'structure') provides a basic description of a dataset.<br><br>

```{r tut=TRUE}
# STATA describe
str(airquality)


```
<br>Return to [index](#index)<br>


# dotplot {#dotplot}

<tt>dotchart</tt> provides a basic desription of a dataset.<br><br>

```{r tut=TRUE}
# STATA Temp, over(Month)
dotchart(airquality$Temp, group=airquality$Month)


```

<br><br>Return to [index](#index)<br>

# drop keep {#drop}

The easiest way is to use a conditional index in square brackets to keep some variables.
For drop use the function <tt>subset</tt>.

```{r tut=TRUE}
# STATA keep Month Day
keep.data <- airquality[, c("Month", "Day")]
head(keep.data)
# STATA drop Month Day
drop.data <- subset(airquality, select = -c(Month, Day))
head(drop.data)


```

<br><br>Return to [index](#index)<br>

# egen {#egen}

Stata's <tt>egen</tt> command is extremely felxible and powerful. 
Some typical examples are provided below.<br> 
<!-- Example with egen by group:<br> -->
<!-- <tt> -->
<!-- &#35; STATA bysort case: egen  meanage = mean(age)<br>  -->
<!-- library(dplyr) <br>  -->
<!-- infert$meanage <- infert %>% <br>  -->
<!--         group_by(case) %>%<br>  -->
<!--         mutate(meanage = mean(age, na.rm=T)) %>%<br>  -->
<!--         pull()<br>  -->
<!-- head(infert)<br>  -->
<!-- </tt> -->
<!-- <br><br>  -->

```{r tut=TRUE}
# STATA egen rankage = rank(age)
infert$rankage <- rank(infert$age)
# STATA egen Nabort = rowtotal(spontaneous induced)
infert$Nabort <- rowSums(infert[, c("spontaneous","induced")], na.rm = T)
# STATA egen excluded = rowmiss(_all)
airquality$excluded <- rowSums(is.na(airquality))
table(airquality$excluded)

#STATA bysort case: egen  meanage = mean(age)<br> 
library(dplyr) 
airquality <- airquality %>%  
        group_by(Month) %>% 
        mutate(mTemp = mean(Temp, na.rm=T)) 
table(airquality$mTemp)


```
<br>Return to [index](#index)<br>

# estimates store lrtest {#lrtest}

Results from statistical models can be stroired by simply assign them to a new R object
<tt>mymodel <- lm(a ~ b, data = c)</tt>. Comparing 2 models with the likelihood ratio test 
use  <tt>anova</tt>.<br><br> 

```{r tut=TRUE}
# STATA regress Ozone Wind Temp
#       estimates store fullmod
#       regress Ozone
#       estimates store nullmod
#       lrtest fullmod nullmod
fullmod <- lm(Ozone ~ Wind + Temp, data=airquality)
nullmod <- lm(Ozone ~ 1, data=airquality)
anova(fullmod, nullmod)

```
<br>Return to [index](#index)<br>


# generate {#generate}

A new variable in R is simply generated by assigning new values to <tt>mydata$newvarname</tt><br><br>

```{r tut=TRUE}
# STATA genrate Celsius = (Temp - 32) * 5/9
airquality$Celsius <- (airquality$Temp - 32) * 5/9
summary(airquality$Celsius)

```
<br>Return to [index](#index)<br>

# graph bar {#bar} 
R's <tt>barplot</tt> function can be used with a numbers or a matrix.
To generate the latter one <tt>prop.table</tt> may be useful.
<br><br>


```{r tut=TRUE}
# STATA graph bar (mean) Ozone, over(Month)
Oz.mean <- by(airquality$Ozone, airquality$Month, mean, na.rm = T)
barplot(Oz.mean)

# STATA's user written command catplot
barplot(prop.table(table(infert$parity, infert$education), margin=2))

```
<br>Return to [index](#index)<br>

# graph box {#box} 
In R: <tt>boxplot</tt>.
<br><br>

```{r tut=TRUE}
# STATA graph box Ozone, over(Month)
boxplot(Ozone ~ Month, data=airquality)

```
<br>Return to [index](#index)<br>

# graph matrix {#matrix} 
<tt>plot(x)</tt> is a generic function. If <tt>x</tt> is a dataset, R will produce a scatterplot matrix. <tt>pairs</tt> is an alternative function which offers more flexibility.
<br><br>
Examples:<br><br>
<tt>
plot(airquality)<br>
pairs(airquality, upper.panel = NULL, col=airquality$Month)
</tt>

<br><br>>Return to [index](#index)<br>

# help findit {#help}

The basic function to access the help documentation is simply a <tt>?</tt> followed by the function.
A search in the documentation can be done with <tt>??</tt> or <tt>help.search</tt><br>
Examples:<br><br>
<tt>
?airquality<br>
??lm<br>
help.search("linear models")
</tt>

<br><br>Return to [index](#index)<br>


# histogram {#hist}

In R: <tt>hist</tt>.
<br><br>

```{r tut=TRUE}
# STATA hist Ozone, start(0) width(10)
hist(airquality$Ozone, breaks=seq(0,170,10))

```
<br>Return to [index](#index)<br>


# if {#if} 
We refer here to Stata's <tt>command <b>if</b> expression</tt>. Not to the programming command.
<br><br>

```{r tut=TRUE}
# STATA summarize if Month == 6
summary(airquality[airquality$Ozone == 6, ])
# STATA summarize Wind if Month == 6
summary(airquality$Wind[airquality$Ozone == 6])
# STATA summarize if Ozone == . 
summary(airquality[is.na(airquality$Ozone), ])
# STATA summarize if Ozone <= 30
# in R better %in% instead of == because of NA
summary(airquality[airquality$Ozone %in% 0:30, ])


```

<br><br>Return to [index](#index)<br>


# in {#in} 
In R usually done with square brackets.
<br><br>

```{r tut=TRUE}
# STATA summarize in 1/10 Month == 6
summary(airquality[1:10, ])


```

<br><br>Return to [index](#index)<br>


# kappa {#kappa} 
The basic R statistic package has no fucntion to calculate the Kappa statistic.
The packages <tt>psych</tt> and  <tt>irr</tt> provide functions to calculate Kappa
and various other types of rater agreement. 
Note: the interpretation of Kappa is not as straight forward as widely assumed. See
https://en.wikipedia.org/wiki/Cohen%27s_kappa#Limitations
<br><br>

<br><br>Return to [index](#index)<br>



# levelsof {#levelsof}

In R simply: <tt>levels</tt>.
<br><br>

```{r tut=TRUE}
# STATA levelsof education  
levels(infert$education)

```
<br>Return to [index](#index)<br>

# list {#list}

To print record numbers 4 to 6 use <tt>mydata[4:6, ]</tt>.
To print records which fulfill a certain condition use <tt>mydata[mycondition, ]</tt>.
<br><br>

```{r tut=TRUE}
# STATA list in 4/6
airquality[4:6, ]
# STATA list of Ozone == 0
airquality[airquality$Solar.R < 10, ]
# to get rid of missing:
airquality[airquality$Solar.R  %in% 0:10, ]

```
<br>Return to [index](#index)<br>

# logistic logit {#logit}

In R, the logistic regression is considered a special case of a broader group of statistical 
models known as generelaized linear models.
The function to fit those models is <tt>glm</tt> (for 'generalized linear model').
<br><br>

```{r tut=TRUE}
# generate binary variable abort = TRUE if induced > 0 or spontaneous > 0   
# STATA generate abort = induced > 0 | spontaneous > 0  
infert$abort <- infert$induced > 0 | infert$spontaneous > 0
# crosstab 
# STATA tab abort case
table(infert$abort, infert$case)
# STATA logit case abort, OR 
model.1 <- glm(case ~ abort, data=infert, family="binomial") 
summary(model.1)
# Get odds ratio
exp(coef(model.1))
# Get confidence intervals
exp(confint(model.1))

```
<br>Return to [index](#index)<br>

# lowess lfit {#lowess}

To generate a scatter plot with a smoothing or regression line you first have to generate the
plot. Afterwards you can add the lines. Note on lowess smoother: if you have missing data in
one of the variables you should first remove those observations (or use the alternative
<tt>loess</tt> which is slightly more complicated to handle). 
<br><br>

```{r tut=TRUE}
# STATA lowess Wind Temp
plot(airquality$Wind, airquality$Temp)
lines(lowess(airquality$Wind, airquality$Temp))


# STATA graph twoway (lfit Wind Ozone) (scatter Wind Ozone )
plot(airquality$Wind, airquality$Temp)
abline(lm(Temp ~ Wind, data = airquality))


```
<br>Return to [index](#index)<br>


# mean {#mean}

Stata's mean calculates the mean together with standard error and confidence intervals.
In R, confidence intervals can be estimated with a one sample t-test
and the (constant only) linear regression returns the standard error.
For geometric and harmonic means see [ameans](#ameans)<br><br>

```{r tut=TRUE}
# STATA mean Temp
# 
# 1 sample t.test for mean and 95%CI 
t.test(airquality$Temp)
#
# linear regression for mean and std Err
summary(lm(Temp ~ 1, data=airquality))


```
<br>Return to [index](#index)<br>

# merge {#merge}

R has a basic <tt>merge</tt> function but it is recommended to use the <tt>dplyr</tt> package.
Have a look at the documentation:<br>
<tt>?dplyr::join</tt>
<br><br>

<br>Return to [index](#index)<br>

# mhodds {#mhodds}

The Mantel-Haenszel Test is implemented in <tt>mantelhaen.test</tt><br><br>

```{r tut=TRUE}
# genrate binary variable abort = TRUE if induced > 0 or spontaneous > 0   
# STATA generate abort = induced > 0 | spontaneous > 0  
infert$abort <- infert$induced > 0 | infert$spontaneous > 0
# crosstab 
# STATA tab abort case
table(infert$abort, infert$case)
# Mantel-Haenszel Test stratified by educational level
# STATA mhodds case abort eduction
mantelhaen.test(infert$case, infert$abort, infert$education)


```
<br>Return to [index](#index)<br>

# mvdecode mvencode {#mvdecode}

Handling missing data in R is a bit tricky. The function <tt>is.na</tt> checks if a value 
is missing. <br><br>

```{r tut=TRUE}
# STATA mvdecode Month, mv(9)  
airquality$Month[airquality$Month == 9] <- NA
table( airquality$Month, useNA="ifany")
# STATA mvdecode _all, mv(999)  
airquality[airquality == 999] <- NA

# STATA mvencode Ozone, mv(999)  
airquality$Ozone[is.na(airquality$Ozone)] <- 999
table( airquality$Ozone, useNA="ifany")


```
<br>Return to [index](#index)<br>

# predict {#predict}

Predicted values in regression models can be estimated via <tt>predict</tt>. 
Residuals are stored in <tt>mymodel$residuals</tt>.
</tt><br><br>

```{r tut=TRUE}
# Predicted values with no missing data
# STATA regress Wind Temp
# STATA predict pred, xb 
model.1 <- lm(Wind ~ Temp, data=airquality) 
summary(model.1)
airquality$pred <- predict(model.1)
# STATA predict resid, residuals
airquality$resid <- model.1$residuals
summary(airquality)

# Predicted values with missing data
model.2 <- lm(Ozone ~ Solar.R, data=airquality) 
summary(model.2)
airquality$pred <- predict(model.2, newdata = airquality)
airquality$resid[-c(model.2$na.action)] <- model.2$residuals
summary(airquality)

```
<br>Return to [index](#index)<br>

# qnorm {#qnorm}

Use <tt>qqnorm</tt> to genrate a normal Q-Q plot. 
See [regress](#regress) for a normal Q-Q plot of the residuals from a regression model. 
<br><br>

```{r tut=TRUE}
# STATA qqnorm Wind
qqnorm(airquality$Wind)


```
<br>Return to [index](#index)<br>

# ranksum signrank ttest {#ttest}

R's function for Student's t-test is <tt>t.test</tt>. 
Note that Stata assumes by default equal variances whereas R assumes unequal variances.
Stata's <tt>ranksum</tt> and <tt>signrank</tt> translates to <tt>wilcox.test</tt> in R. 
<br><br>

```{r tut=TRUE}
# unpaired t-test (equal var)
# STATA ttest age, by(case)
t.test(age ~ case, data = infert, var.equal = T)
# unpaired t-test (unequal var - Welch's approx)
# STATA ttest age, by(case) welch
t.test(age ~ case, data = infert)
# nonparametric
# STATA ranksum age, by(case) welch
wilcox.test(age ~ case, data = infert)

# paired t-test
# STATA ttest spontabeous == induced
t.test(infert$spontaneous, infert$induced, paired=T)
# nonparametric Wilcoxon matched-pairs signed-ranks test
# STATA signrank spontabeous == induced
wilcox.test(infert$spontaneous, infert$induced, paired=T)


```
<br>Return to [index](#index)<br>


# recode {#recode}

There is no function in R which has the same flexibility as Stata's <tt>recode</tt>.
A useful function is <tt>recode</tt> in the <tt>dplyr</tt> package.
Note that numeric values have to be enclosed in backticks ` and that replacements have to be specified exhaustively (other values will be set to <tt>NA</tt>).
If you would like to recode only a single or few values see <tt>[replace](#replace)</tt>.<br>
Stata has also a <tt>recode</tt> function which can be used together with 
<tt>generate</tt> to split numerical values into groups.
If you would like to split a numerical variable into quantiles see <tt>[xtile](#xtile)</tt>
<br><br>

```{r tut=TRUE}
# Recoding numerical values
# STATA recode Month (5 9 = 0) (6/8 = 1), generate(Summer) 
library(dplyr)
airquality$Summer <- recode(airquality$Month, `5` = 0, `6` = 1, `7` = 1, `8` = 1, `9` = 0)
airquality$Summer

# Categrize Ozone into <=50, <=100, 101+
# STATA generate OzoneCat = recode(Ozone, 50, 100, 200) 
airquality$OzoneCat <- cut(airquality$Ozone, breaks=c(0,50,100,200), include.lowest=T)
airquality$OzoneCat

```
<br>Return to [index](#index)<br>

# regress  rvfplot {#regress}

R's function for linear regression is <tt>lm</tt> (for 'linear model').
Regression diagnostic plots can be obtained with the generic <tt>plot</tt> function.
<br><br>

```{r tut=TRUE}
# STATA regress Temp Ozone 
model.1 <- lm(Temp ~ Ozone, data=airquality) 
summary(model.1)

# Get confidence intervals
confint(model.1)

# Ressiduals vs fitted plot
plot(model.1, which = 1)

# normal Q-Q plot of the residuals
plot(model.1, which = 2)

```
<br>Return to [index](#index)<br>

# rename {#rename}

Variable names can be manipulated with <tt>colnames(mydata)</tt>.
<br><br>

```{r tut=TRUE}
# STATA rename Solar.R solar
colnames(airquality)
colnames(airquality)[2] <- "solar"
colnames(airquality)
# STATA rename _all, lower
colnames(airquality) <- tolower(colnames(airquality))
colnames(airquality)

```
<br>Return to [index](#index)<br>

# replace {#replace}

<tt>replace</tt> is usually used with the conditional statement <tt>if</tt>.
In R this is either done via conditional indexing, the <tt>ifelse</tt> function or 
<tt>mutate</tt> in the <tt>dplyr</tt> package.<br><br>

```{r tut=TRUE}
# with indexing
# STATA replace age = 40 if age > 40
infert$age[infert$age > 40] <- 40
table(infert$age)
# with ifelse
# STATA replace age = 35 if age > 35
infert$age <- ifelse(infert$age > 35, 35, infert$age)
table(infert$age)
# with mutate
# STATA replace age = 30 if age > 30
library(dplyr)
infert <- mutate(infert, age = ifelse(age > 30, 30, age))
table(infert$age)

```
<br>Return to [index](#index)<br>

# reshape {#reshape}

A function with the same name is available in R. If you often transform data, 
consider the package <tt>reshape</tt>.<br><br>

```{r tut=TRUE}
# with indexing
# STATA reshape wide Ozone Solar.R Wind Temp, i(Day) j(Month)
air.wide <- reshape(airquality, direction="wide",idvar="Day",timevar="Month")
head(air.wide)


```
<br>Return to [index](#index)<br>


# sample {#sample}

A sub sample of the dataset can be obtained with the <tt>sample</tt> function but the
<tt>dplyr</tt> package provides more features including stratified sampling. 
<br><br>

```{r tut=TRUE}
# STATA sample 4, count by(Month) 
library(dplyr)
sub.sample <- airquality %>%
  group_by(Month) %>%
  sample_n(4)
head(sub.sample, 10)


```
<br>Return to [index](#index)<br>


# save {#save}

To save a dataset in a separated format use the function 
<tt>write.table(mydata, "mydata.txt")</tt>. 
There are in addition the functions <tt>write.csv</tt> and <tt>write.csv2</tt> which are
doing the same with different default settings for the arguments. However, it
is recommended to specify the arguments individually (see the example).
Note: R's <tt>save</tt> function is doing something else. 
It saves one or several R objects to an Rdata file. 
<br>Examples:<br><tt>
write.table(infert, "mydata.csv", sep= "&#92;t",  na= "NA", dec= ".", row.names= F, col.names = T)<br>
&#35; Example below does only work on Windows:<br>
write.table(infert, "clipboard", sep= "&#92;t",  na= "NA", dec= ".", row.names= F, col.names= T)
</tt>
<br><br>


<br>Return to [index](#index)<br>


# scatter {#scatter}

R's generic <tt>plot</tt> function will produce a scatter-plot if the first 2 arguments 
are numeric variables. 
<br><br>

```{r tut=TRUE}
# STATA scatter Ozone Temp 
plot(airquality$Ozone, airquality$Temp)

# STATA twoway (scatter Ozone Temp if Month == 5) ...  (scatter Ozone Temp if Month == 9) 
plot(airquality$Ozone, airquality$Temp, col=airquality$Month)


```
<br>Return to [index](#index)<br>



# sort {#sort}


Sorting a dataset in R is not really intuitive. Therefore, many people prefer the R package <tt>dplyr</tt><br><br>

```{r tut=TRUE}
# STATA sort Temp 
airquality <- airquality[order(airquality$Temp), ]
head(airquality)
# using package dplyr
library(dplyr)
airquality <- arrange(airquality, Temp)
head(airquality)

```
<br>Return to [index](#index)<br>

# tabulate tab1 {#tabulate}

Useful functions in R include <tt>table prop.table addmargins</tt>.
Note: R will omit missing values unless the argument <tt>useNA="ifany"</tt> is specified. Unlike Stata, Chi^2^ and Fisher's exact test have to be specified. <br><br> 

```{r tut=TRUE}
# STATA tabulate Ozone
table(airquality$Ozon)
# STATA tabulate Ozone, missing
table(airquality$Ozon, useNA="ifany")

# genrate binary variable abort = TRUE if induced > 0 or spontaneous > 0   
# STATA generate abort = induced > 0 | spontaneous > 0  
infert$abort <- infert$induced > 0 | infert$spontaneous > 0
# STATA tabulate abort case
table(infert$abort, infert$case)
addmargins(table(infert$abort, infert$case))
# STATA tabulate abort case, row
prop.table(table(infert$abort, infert$case), margin = 1)
# STATA tabulate abort case, chi exact
chisq.test(infert$abort, infert$case)
fisher.test(infert$abort, infert$case)

# STATA tab1 _all
apply(infert, 2, table)

```
<br>Return to [index](#index)<br>


# xi {#xi}

Stata's xi prefix can be used for different purposes. 
Particular important is that the <tt>xi: ... i.myvariable </tt> notation to tell Stata 
that <tt>myvariable</tt> is a categorical variable. Categorical variables are called 
factors in R and the first step when we use a new dataset should be to explore which 
variables are numeric and which categorical using the <tt>str</tt>. A numeric variable 
can be converted to a factor using the function <tt>as.factor</tt>.


```{r tut=TRUE}
# Inspect the variable types of data infert
str(infert)
# Logistic regression to assess the effect of education on case
# Note: this variable is a factor
# STATA xi: logit case i.eduction
model.1 <- glm(case ~ education, data=infert, family="binomial")
summary(model.1)
# Logistic regression to assess the effect of induced on case
# Note: this variable is numeric
# STATA xi: logit case i.induced
model.2 <- glm(case ~ as.factor(induced), data=infert, family="binomial")
summary(model.2)
```
<br>Return to [index](#index)<br>

# xtile {#xtile}

To split a numeric variable into quantiles you need to combine <tt>cut</tt> and <tt>quantile</tt>. alternatively you can use <tt>ntile</tt> (<tt>dplyr</tt> package).<br><br> 

```{r tut=TRUE}
# split Ozone into quartiles
# STATA xtile Ozone4 = Ozone, nq(4)
airquality$Ozone4 <- cut(airquality$Ozone, 
          breaks = quantile(airquality$Ozone, probs=c(0, 0.25, 0.5, 0.75, 1), na.rm = T),
          include.lowest = T)
table(airquality$Ozone4)
#
# split Temp into quartiles using the dplyr package
# STATA xtile Temp4 = Temp, nq(4)
library(dplyr)
airquality <- mutate(airquality, Temp4 = ntile(Temp, 4))
table(airquality$Temp4)
```
<br>Return to [index](#index)<br>
