---
title: "Writing own functions in R"
output:
  bookdown::html_document2: 
    number_sections: false 
    toc: true
    toc_float: true
    smart: false
---
  
  
```{r, include=FALSE}
knitr::opts_chunk$set(comment = NA)
tutorial::go_interactive(greedy=FALSE,  height = 50)
```

# Introduction {#intro}

One strength of R is the user's ability to define own functions. 
A function is simply some R code to perform a specific task.
They are especially useful of you have some R code which you want to use repeatedly.
You don't need functions, every task can also be done with plain code but functions have 
several advantages:<br><br>
Your code might be shorter and easier to interpret<br> 
Your code is easier to maintain<br>
It is easier to reuse functions in future R scripts<br><br> 

You can write your own functions with the following syntax:<br><br> 
<tt>my_function <- function(my_arguments) {</tt><br> 
    <tt> # some R code to be executed </tt><br> 
    <tt> return(my_result)</tt><br>
<tt>}</tt><br><br>

We define the function <tt>my_function</tt> with the keyword <tt>function()</tt>.
The names of one and more arguments can be specified in round parentheses.
Most functions require at least one argument but even if a function does not 
require any arguments (e.g. <tt>dir()</tt> you have to type the round parentheses.
Next, the body of the function – the R code to be executed – is contained within curly parentheses.
Finally, you specify which value (i.e. which result) should be returned if the function is called with <tt> return(my_result)</tt>. 
If you don't explicitly define a return value, the function will return 
the last expression evaluated within the within curly parentheses.<br><br>

 Traditionally, the first program written by people learning a new programming language,
 is the famous Hello-World-Program. A program which displays the message "Hello, World!".
 This function does not require any arguments, 
 because independent from the input the output is always the same text message.
 We define a new function with the name <tt>hello</tt> and in the function body 
 we simply write some code to print "Hello, World!" in the R console.<br><br>
 
```{r tut=T, height=20}
# Generate the deck
hello <- function() {
   print("Hello, World!")
}

hello()

```
 
<br>
You can see several things:<br><br>
Although we don't use any arguments the round parentheses are provided after <tt>function</tt>.<br>
Although R usually print all values (in this cases the text "Hello, World!")
automatically if they are not assigned to a new object. 
We used in this example the <tt>print</tt> function because R will not print an
unassigned value if the code is enclosed within curly parentheses.<br>
The function is called by typing the name of the function followed by parentheses.
If you omit the parentheses <tt>hello</tt> R will return the code stored in the function.
<br><br>

# A simple function {#gm}

R has not a function to calculate the geometric mean, so we will define one. 
The geometric mean is the nth root of the product of n numbers.
However, the product can be extremely large if n is getting high it is usually 
calculated as the exponential of the arithmetic mean of logarithms which is in R language:<br>
<tt>exp(mean(log(x)))</tt><br>
where x is a numeric vector. If you don't know the term numeric vector, 
it is simply a collection of numbers, e.g. <tt>c(2, 3.1, 0)</tt>  or <tt>5:10</tt>.<br>
Have a look at the code below ro see how the function is defined:<br><br>

```{r tut=T, height=50}
geom.mean <- function(x) {
   result <- exp(mean(log(x)))
   return(result)
}
some_numbers <- c(1,10,100,1000)
geom.mean(some_numbers)

```

We call the function <tt>geom.mean</tt> together with the argument x. 
x contains the numbers to calculate the geometric mean from.
You could also define the numbers directly in the function call:<br>
<tt>geom.mean(c(1,10,100,1000))</tt><br>
but usually you will apply the function to a variables in a dataset, 
e.g. the variable <tt>Wind</tt> in the dataset <tt>airquality</tt> which is included in R. 
Copy the line below in the code window above and execute the code again.<br>
<tt>geom.mean(airquality$Wind)</tt><br>
The code behaves in the same way as plain R code, 
e.g. if x contains any negative numbers R would show the warning message:<br>
<tt>[1] NaN</tt><br>
<tt>Warning message:</tt><br>
<tt>In log(x) : NaNs produced</tt><br>
<br>
Of course, you can assign the result to a new R object for later use 
in the same ways you do it with other functions:<br>
<tt>gm1 <- geom.mean(c(1,10,100,1000))</tt><br>
<br><br>

<b>Exercise</b><br> 
The variable  <tt>Ozone</tt> in the dataset <tt>airquality</tt> has several 
missing observations.
Therefore, the following line:<br> 
<tt>geom.mean(airquality$Ozone)</tt><br>
will return <tt>NA</tt>.<br>
Can you modify the code within the curly parentheses to omit the missing 
values before the geometric mean is calculated (<tt>na.rm = T</tt>)?<br><br> 

# Multiple arguments {#args}

In most functions you can specify more than one argument. 
The code below defines a new function <tt>exponentiate</tt> which has to arguments:
the base b and the exponent n.
<br><br>

```{r tut=T, height=50}
exponentiate <- function(b, n = 2) {
   result <- b^n
   return(result)
}

exponentiate(2, 10)
```

If you have multiple arguments, R expects that you provide them in the same order as 
defined in the syntax. 
If you would like to specify them in a different order, you have to provide in addition the argument names, i.e.:
<tt>exponentiate(n = 10, b = 2)</tt><br>
You have likly noticed that the argument n has a default value (<tt>b = 2</tt>).
If you omit the argument, R will use the default value, e.g.:
<tt>exponentiate(7) # result will be 7^2 = 49</tt><br>
<br><br>

<b>Exercise: Ancient Babylonian multiplication</b><br> 
In ancient Babylon they had no positional numeral system (place-value notation).
Like in ancient Rome they used symbols, each with a fixed integer value (like 1, 10, 100).
Multiplication is not straight forward, so they used the following formula to multiply 2 numbers (x * y):<br>
<tt>((x + y)^2 - (x - y)^2) / 4</tt><br>  
Complete the code below to define the function babylon.<br><br>

```{r ex="babylon", tut=T, type="sample-code", height=50}
babylon <- function(x, y) {
   # type the code for the Babylonian multiplication here 
   return(result)
}

babylon(66, 55)
```

```{r ex="babylon", tut=T, type="solution", height=50}
babylon <- function(x, y) {
    result <- ((x + y)^2 - (x - y)^2) / 4
   return(result)
}

babylon(66, 55)
```
<br><br>

# Multiple results {#res}

If you would like to return more than 1 result, the best way is to return them as a list.
If you don't know what a list is: it is a collection of items like a vector but the
single items can be of different classes, e.g. numbers, text, data frames, etc.
The elements are usually named and can be accessed 
like a variable in a data frame with the $ sign.<br>
Let us define a function which calculates the odds ratio of a 2x2 table and
the confidence interval. The upper limit of the confidence interval can be 
approximated with the follwiong formula:<br>
<tt>OR + 1.96 * sqrt(1/A + 1/B + 1/C + 1/D)</tt><br><br>  

```{r tut=T, height=50}
# A = number of exposed with disease, .., D = number unexposed without disease
OR <- function(A, B, C, D) {
   OR <- (A * D) / (B * C)
   conf.int.low <- OR - 1.96 * sqrt(1/A + 1/B + 1/C + 1/D)
   conf.int.high <- OR + 1.96 * sqrt(1/A + 1/B + 1/C + 1/D)
   return(list(OR = OR, conf.int = c(conf.int.low, conf.int.high)))
}

or1 <- OR(30, 70, 40, 80)
or1
# Return only the confidence interval
or1$conf.int
```

<br><br>

# Condiotional execution with <tt>if</tt> {#if}

Let us revisit the geometric mean example. The geometric mean is often used for 
parasitic egg counts in stool samples. However, there is one problem: there are 
usually also some negative stool samples, i.e. stool samples with 0 eggs. 
If there is at least 0, the geometric mean is 0. Therefore, the formula is usually
modified as:<br>
<tt>exp(mean(log(x + 1))) - 1</tt><br>
Our aim is to write a function which calculates the modified geometric mean if there 
is at lest one 0 in x and the original formula if all numbers are positive.
To check if there is at least one 0, we van use: <br> 
<br>
<tt>if (any(x == 0)){</tt><br>
<tt>  # code to be executed</tt><br>
<tt>}</tt><br><br>

After <tt>if</tt>a condition is provided in round parentheses and only if the condition 
is <tt>TRUE</tt> the code in the curly parentheses will be executed. As you see, it
follows more or less the same logic as the parentheses for the functions.
<br><br>


```{r ex="gm2", tut=T, type="sample-code", height=50}
geom.mean <- function(x) {
  if(any(x == 0)) { 
       print("0 detected, +1 added")
       result <- exp(mean(log(x + 1))) - 1
  }
  if(all(x > 0)) { 
       result <- exp(mean(log(x)))
  }  
  return(result)
}

geom.mean(c(0,10,100,1000))
```

```{r ex="gm2", tut=T, type="solution", height=50}
geom.mean <- function(x) {
  if(any(x < 0)) { 
       stop("Negative numbers")
  }  
  if(any(x == 0)) { 
       print("0 detected, +1 added")
       result <- exp(mean(log(x + 1))) - 1
  }
  if(all(x > 0)) { 
       result <- exp(mean(log(x)))
  }  
  return(result)
}

geom.mean(c(-1,10,100,1000))
```

<br><br>

<b>Exercise</b><br> 
It is impossible to calculate Geometric Mean with negative numbers. 
Modify the code above that it produces an error with a warning massage 
<tt> stop("Negative numbers")</tt> if there are any negative numbers in x.<br><br>

# Anonymous functions  {#anon}

Sometimes you need just a very short function and you need it only once. 
This is for instance sometimes the case if you have your own function within the    
<tt>by</tt> function. It is possible to define such a function directly without
proper definition and without an own name.
In the example below you would like to calculate the mean and the standard deviation of 
the variable <tt>Wind</tt> stratified by month (data </tt>airquality</tt>).<br>
Remember: <tt>by</tt> is used in the following way:<br>
<tt> by(object_to_analyse, vector_defining_strata, function_to_apply)</tt><br><br>
You need only a single line!<br><br>


```{r tut=T, height=50}
by(airquality$Wind, airquality$Month, function(x) c(mean(x), sd(x)))

```

# Passing arguments with <tt>...</tt> {#dots}

Often a function is called within another function, e.g. in our geometric mean example
we call the built-in function <tt>mean</tt> within our own <tt>geom.mean</tt> function.
Sometimes you wish that some arguments of your newly defined function
should be forwarded to a function called within it. 
If you know in advance which arguments you would like to forward, you can simply
put them as additional arguments. However, sometimes you do not know exactly at the time 
of writing the code which additional arguments could be useful.
In such a case you can simply put a placeholder in the list of arguments which is done 
with <tt>...</tt>.<br><br>
In our initial geometric mean example we had some difficulties with missing values.
We could expand the arguments with the 2 dots as follows:<br><br>

```{r tut=T, height=50}
geom.mean <- function(x, ...) {
   result <- exp(mean(log(x), ...))
   return(result)
}
some_numbers <- c(NA,10,100,1000)
geom.mean(some_numbers)
geom.mean(some_numbers, na.rm = T)
```


